Dynamics of Colombo's Top: Generating Exoplanet Obliquties from Planet-Disc Interactions

### 2020-04-15 writeups/Su_disk_disp.tex
another nonad thetasdi writeup

It is natural to imagine some of the mechanisms that generate planetary
obliquities in the Solar System may also operate in exoplanetary systems.
Recently, \citet{millholland_disk} studied the production of planet obliquities
via spin-orbit resonance, where a dissipating protoplanetary disk causes
resonance capture and advection. In their work, a planet is accompanied by an
inclined exterior disk; as the disk graduatelly dissipates, the resulting
planetary obliquity is generally substantial, reaching $90^\circ$ for what they
characterize as adiabatic resonance crossings.

@@

The paper is organized as follows. In Section~\ref{s:eq}, we review the relevant
spin-orbit dynamics, equations and key concepts that are used in the remainder
of the paper. In Sections~\ref{s:ad} and~\ref{s:nonad}, we study the evolution
of the system when the disk dissipates on different timescales, from highly
adiabatic to nonadiabatic. Analytical results are presented to explain the
numerical results in both limits. We discuss the implications of our results in
Section~\ref{s:disc}. Our primary physical results consist of
Fig.~\ref{fig:ad_ensemble} in the adiabatic limit and
Fig.~\ref{fig:3_ensemble_05_15} in the nonadiabatic limit. Some detailed
calculations are relegated to the appendicies, including a leading-order
estimate of the final planetary obliquities given small initial spin-disk
misalignment angles in Appendix~\ref{s:ad_approx}.

@@

Spin states satisfying $\p{d\uv{s}/d\uv{\tau}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs) \citep{colombo1966,peale1969}. They require that
$\uv{s}$, $\uv{l}$, and $\uv{l}_{\rm d}$ be coplanar. There are either two or
four CSs, depending on the value of $\eta$. They are specified by the obliquity
$\theta$ and the precessional phase of $\uv{s}$ around $\uv{l}$, denoted by
$\phi$. Following the standard convention and nomenclature (see
Figs.~\ref{fig:cs_vecs} and~\ref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0$ and
$\theta < 0$, corresponding to $\uv{s}$ and $\uv{l}_{\rm d}$ being on opposite
sides of $\uv{l}$, while CS2 has $\phi = \pi$ and $\theta > 0$, corresponding to
$\uv{s}$ and $\uv{l}_{\rm d}$ being on the same side of $\uv{l}$. When $\eta <
\eta_{\rm
c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2},\label{eq:etac}
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.

@@

Since $\p{\phi, \cos \theta}$ are canonically conjugate, the integral $\oint
\cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant (see
Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically
\citep{henrard1987,ward2004I}. Define
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1}.
    \end{align}
\end{subequations}
The areas for $\eta < \eta_{\rm c}$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{\rm I} &= 2\pi\p{1 - z_0} - \frac{A_2}{2}, \label{eq:A1}\\
        A_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},
            \label{eq:A2}\\
        A_{\rm III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. While the
zones are not formally defined for $\eta > \eta_{\rm c}$ since the separatrix
disappears, a natural extension exists: evolve an initial phase space point $p$
under adiabatic decrease of $\eta$ until the separatrix appears at $\eta =
\eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm c}$. Since
phase space area is conserved under adiabatic evolution, this extension implies
$A_{\rm i}\p{\eta > \eta_{\rm c}} = A_{\rm i}(\eta_{\rm c})$. The boundary
between these extended zones is denoted by the dashed black line in panel (a) of
Fig.~\ref{fig:eq_1contours}, where no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/1_areas.png}
    \caption{The solid lines show the fractional areas of each of the zones
    $A_{\rm i}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward}. The colored
    dashed lines correspond to small $\eta$ approximations used in
    Appendix~\ref{s:ad_approx}. The colored dashed lines for $\eta > \eta_{\rm
    c}$ are the effective values of $A_{\rm II}, A_{\rm III}$ for $\eta >
    \eta_{\rm c}$, denoting the points that would flow into either area under
    adiabatic decrease of $\eta$ from $\eta > \eta_{\rm c}$ (see the text). The vertical black
    dashed lines correspond to $\eta = \eta_{\rm c}$ [Eq.\eqref{eq:etac}] and
    the values of $\eta$ for which $A_{\rm II}$ is maximized ($\eta_{\rm \max,
    II}$) and for which $A_{\rm III}$ is minimized ($\eta_{\rm \min, III}$,
    Eq.~\eqref{eq:eta_minIII}).}\label{fig:eq_areas}
\end{figure}

@@

In this section, we study the evolution of the planetary obliquity $\theta$ when
the parameter $\eta$ [or the disk mass $M_{\rm d}$; see Eqs.~\eqref{eq:eta}
and~\eqref{eq:deta_dt}] decreases sufficiently slowly that the evolution is
adiabatic. Strictly speaking, this requires the disk evolution timescale $t_{\rm
d}$ [Eq.~\eqref{eq:dmd_dt}] be much larger than all timesccales of the dynamical
system governed by the Hamiltonian [Eq.~\eqref{eq:H}]. This is of course not
possible in all cases, as the motion along the separatrix has an infinite
period.

@@

We must next understand the above results in terms of the initial spin-disk
misalignment angle $\theta_{\rm sd, i}$. To do so, we first compute the final
obliquity when $\theta_{\rm sd, i} = 0$, corresponding to initial obliquity
$\theta_{\rm i} = I$. The set of possible initial conditions
for nonzero $\theta_{\rm sd, i}$ is a ring of angular width $\theta_{\rm sd, i}$
about $\uv{l}_{\rm d}$. We find this ring evolves to a new ring of angular width
$\theta_{\rm sd, i}$ about the $\theta_{\rm sd, i} = 0$ prediction, so the range
of final obliquities is the $\theta$ range spanned by this final
ring\footnote{The interpretation for this procedure is that the process
generating the final obliquity predicted by Eq.~\eqref{eq:nonad_q_f} acts only
on the \emph{center} of libration of the trajectory when $\eta \gg 1$ (CS2).
This is not apparent from our derivation but reproduces the results of the
numerical simulation.}. This procedure yields prediction
\begin{equation}
    \abs{\sqrt{\frac{2\pi}{\epsilon}}\frac{\sin \p{2I}}{2}
        - \theta_{\rm sd, i}}
    \leq \theta_{\rm f}
    \leq \sqrt{\frac{2\pi}{\epsilon}}\frac{\sin\p{2I}}{2} + \theta_{\rm sd, i}
    .\label{eq:nonad_q_f_dist}
\end{equation}
The absolute value in the lower bound is a result of the ring of final
conditions crossing the North pole.

In Fig.~\ref{fig:nonad_3_ensemble}, we show the numerical result of $\theta_f$
vs $\theta_{\rm sd,i}$ for $I = 5^\circ$ and $\epsilon = 0.3$. We see that
Eq.~\eqref{eq:nonad_q_f_dist} provides good lower and upper bounds of the final
obliquity for $\theta_{\rm sd,i} \lesssim 45^\circ$.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/3_ensemble_05_05.png}
    \caption{$\theta_{\rm  f}$ vs $\theta_{\rm sd, i}$ for $I=5^\circ$ and
    $\epsilon = 0.3$ (firmly in the nonadiabatic regime). The blue dots
    represent numerical results, and the two red lines show the analytical lower
    and upper limits given by
    Eq.~\eqref{eq:nonad_q_f_dist}.}\label{fig:nonad_3_ensemble}
\end{figure}

Figures~\ref{fig:nonad_3_scan} and~\ref{fig:nonad_3_scan_20} show $\theta_f$ as
function of $\epsilon$ for $\theta_{\rm sd,i} = 0$ and $I = 5$, $20^\circ$. We
see that the agreement between the numerical results and
Eq.~\eqref{eq:nonad_q_f} is excellent, particularly for the $I = 5^\circ$ case.
The agreement when $I = 20^\circ$ is slightly poorer as $\theta_{\rm i}$ is
larger and hence less negligible in Eq.~\eqref{eq:nonad_q_f}. For $\epsilon \ll
\epsilon_{\rm c}$, we find $\theta_f \simeq 90^\circ$, in agreement with the
result of adiabatic evolution (see Fig.~\ref{fig:ad_ensemble}).

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/3scan.png}
    \caption{Final obliquity $\theta_{\rm  f}$ as a function of $\epsilon$ for
    $\theta_{\rm sd,i} = 0$ and $I = 5^\circ$. The shaded area, bordered by the
    black line, corresponds to the adiabatic regime estimated by
    Eq.~\eqref{eq:ad_constr}. The blue dots are numerical results, and the red
    dashed line corresponds to Eq.~\eqref{eq:nonad_q_f}, which is in good
    agreement with numerical results for $\epsilon > \epsilon_{\rm c} \approx
    0.1$ (the nonadiabatic regime); $\theta_{\rm f} \simeq  90^\circ$ in the
    adiabatic regime ($\epsilon\ll \epsilon_{\rm c}$).}\label{fig:nonad_3_scan}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/3scan_20.png}
    \caption{Same as Fig.~\ref{fig:nonad_3_scan} but for $I=20^\circ$. The right
    shaded area is where $\theta_{\rm f}$ predicted by Eq.~\eqref{eq:nonad_q_f}
    violates the assumption $\theta_{\rm f} \gg \theta_{\rm
    i}$.}\label{fig:nonad_3_scan_20}
\end{figure}

@@

In this section, we linearize the equations of motion near each CS\@. We
determine the stability of each CS and the local libration frequencies of the
stable CSs.

\subsection{Canonical Equations of Motion and Solutions}

For analytical work, we adopt spherical coordinate system where $\uv{l} =
\uv{z}$ and $\theta, \phi$ are the polar and azimuthal angle of $\uv{s}$. We
choose $\uv{l}_{\rm z}$ at coordinates $\theta = I, \phi = \pi$ (see
Figs.~\ref{fig:cs_vecs} and~\ref{fig:cs_locs}). Following modern mathematical
convention, we will use
$\theta \in [0, \pi)$% chktex 9
and $\phi \in [0, 2\pi)$.% chktex 9

The equations of motion in coordinates $\p{\phi, \cos \theta}$ follow by
applying Hamilton's equations to the Hamiltonian [Eq.~\eqref{eq:H}]:
\begin{subequations}\label{se:H_eom}
    \begin{align}
        \rd{\phi}{t} = \pd{\mathcal{H}}{(\cos\theta)}
            &= -\cos\theta + \eta\p{\cos I + \sin I \cot \theta \cos \phi},
                \label{seq:H_eom_phi_t}\\
        \rd{(\cos \theta)}{t} = -\pd{\mathcal{H}}{\phi}
            &= -\eta \sin I \sin \theta \sin \phi.
                \label{seq:H_eom_mu_t}
    \end{align}
\end{subequations}
These agree with Eq.~\eqref{eq:dsdt_base}. While these equations are numerically
stiff owing to the $\cot\theta$ term, they are the most intuitive description
for analytical work.

We next develop some approximate forms for the CS obliquities that will be
useful for guiding later discussion. Denote $\theta_{\rm cs}$ the obliquity of
some Cassini State (fixed point of Eqs.~\eqref{se:H_eom}).
\begin{enumerate}
    \item $\eta \ll 1, \cos \theta_{\rm cs} \ll 1$ --- This is the limiting case
        for CS2 and CS4 when $\eta \ll \eta_{\rm c}$. We examine
        Eqs.~\eqref{seq:H_eom_phi_t} and approximate $\cot \theta \approx \cos
        \theta$ to find
        \begin{equation}
            \cos \theta_{\rm cs} \approx \frac{\eta \cos I}
                {1 \mp \eta \sin I}.
        \end{equation}
        The two signs correspond to choices of $\cos \phi$. Each sign
        corresponds to one of CS2 and CS4.

    \item $\sin \theta_{\rm cs} \ll 1$ --- The $\eta \ll 1, \eta \gg 1$ cases
        can be solved together, and are the limiting cases both for CS1 and CS3
        when $\eta \ll \eta_{\rm c}$, and for CS2 and CS3 when $\eta \gg
        \eta_{\rm c}$. It proves easiest to rewrite Eqs.~\eqref{seq:H_eom_phi_t}
        as
        \begin{equation}
            0 = \cos \theta_{\rm cs}\p{-1 + 2\eta
                \frac{\sin \p{I \pm \theta_{\rm cs}}}{\sin \p{2\theta_{\rm
                    cs}}}}.\label{eq:rewritten_dphi}
        \end{equation}
        The $\pm$ choice again comes from choice of $\cos \phi$. Then assuming
        $\theta_{\rm cs}, I \ll 1$, we obtain $\frac{I \pm
        \theta_{\rm cs}}{2\theta_{\rm cs}} = \frac{1}{2\eta}$ or
        \begin{equation}
            \theta_{\rm cs} = \frac{\eta I}{1 \mp \eta}.
        \end{equation}

        In the limit $\eta \ll \eta_{\rm c}$, these two solutions describe CS1
        and CS3, while in the limit $\eta \gg \eta_{\rm c}$ these two solutions
        describe CS2 and CS3.
\end{enumerate}
Again, the $\theta_{\rm cs}$ values derived here do not follow the same
convention as Fig.~\ref{fig:cs_locs} due to our different $\theta, \phi$
convention.

@@

In this section, we will use approximations valid for small $\eta$ to derive
analytic expressions for the final obliquities at small $\theta_{\rm sd, i}$ and
associated probabilities for the II $\to$ I and II $\to$ III tracks that are
accessible.

We first seek a simple parameterization for the separatrix. We first solve for
equilibria of the equation of motion Eq.~\eqref{eq:dsdt_base} to compute the
coordinates for Cassini State 4:
\begin{equation}
    \cos \theta_4 \approx \frac{\mu \cos I}{1 - \eta \sin I}.
\end{equation}
Note that $\phi_4 = 0$. Then, the separatrix is the level curve of the
Hamiltonian intersecting CS4, so it satisfies $\mathcal{H}\p{\phi, \theta_{\rm
sep}(\phi)} = \mathcal{H}\p{\phi_4, \theta_4}$. To leading order in $\eta$, this
has simple expression
\begin{equation}
    \cos \theta_{\rm sep}(\phi) \approx \cos \theta_4 \pm
        \sqrt{2\eta \sin I\p{1 - \cos \phi}}.
\end{equation}
Integration of the phase area enclosed by the two legs of the separatrix then
yields
\begin{equation}
    A_{\rm II}(\eta) \approx 16\sqrt{\eta \sin I}.\label{eq:a_approx}
\end{equation}
This, in conjunction with Eqs.~\eqref{eq:henrard_hop}, is sufficient to compute
$\theta_{\rm f}\p{\theta_{\rm sd, i}}$ for zone II initial conditions.

>>>>

It is natural to imagine some of the mechanisms that generate planetary
obliquities in the Solar System may also operate in exoplanetary systems.
Recently, \citet{millholland_disk} studied the production of planet obliquities
via a spin-orbit resonance, where a dissipating protoplanetary disk causes
resonance capture and advection. In their work, a planet is accompanied by an
inclined exterior disk; as the disk gradually dissipates, the planetary
obliquity increases, reaching $90^\circ$ for what the authors characterize as
adiabatic resonance crossings.

@@

The paper is organized as follows. In Section~\ref{s:eq}, we review the relevant
spin-orbit dynamics and key concepts that are used in the remainder of the
paper. In Sections~\ref{s:ad} and~\ref{s:nonad}, we study the evolution of the
system when the disk dissipates on different timescales, from highly adiabatic
to nonadiabatic. Analytical results are presented to explain the numerical
results in both limits. We discuss the implications of our results in
Section~\ref{s:disc}. Our primary physical results consist of
Fig.~\ref{fig:ad_ensemble} in the adiabatic limit and
Fig.~\ref{fig:3_ensemble_05_15} in the nonadiabatic limit. Some detailed
calculations are relegated to the appendicies, including a leading-order
estimate of the final planetary obliquities given small initial spin-disk
misalignment angles in Appendix~\ref{s:ad_approx}.

@@

Spin states satisfying $\p{d\uv{s}/d\uv{\tau}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs) \citep{colombo1966,peale1969}. They require that
$\uv{s}$, $\uv{l}$, and $\uv{l}_{\rm d}$ be coplanar. There are either two or
four CSs, depending on the value of $\eta$. They are specified by the obliquity
$\theta$ and the precessional phase of $\uv{s}$ around $\uv{l}$, denoted by
$\phi$. Following the standard convention and nomenclature (see
Figs.~\ref{fig:cs_vecs} and~\ref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0$ and
$\theta < 0$, corresponding to $\uv{s}$ and $\uv{l}_{\rm d}$ being on opposite
sides of $\uv{l}$, while CS2 has $\phi = \pi$ and $\theta > 0$, corresponding to
$\uv{s}$ and $\uv{l}_{\rm d}$ being on the same side of $\uv{l}$.
The CS obliquity satisfie
\begin{equation}
    \sin \theta \cos \theta - \eta \sin\p{\theta - I} = 0.
\end{equation}
When $\eta < \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2},\label{eq:etac}
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.

@@

Since $\p{\phi, \cos \theta}$ are canonically conjugate, the integral $\oint
\cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant (see
Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically
\citep{henrard1987,ward2004I}. Define
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1}.
    \end{align}
\end{subequations}
The areas for $\eta < \eta_{\rm c}$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{\rm I} &= 2\pi\p{1 - z_0} - \frac{A_2}{2}, \label{eq:A1}\\
        A_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},
            \label{eq:A2}\\
        A_{\rm III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. While the
zones are not formally defined for $\eta > \eta_{\rm c}$ since the separatrix
disappears, a natural extension exists: evolve an initial phase space point $p$
under adiabatic decrease of $\eta$ until the separatrix appears at $\eta =
\eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm c}$. Since
phase space area is conserved under adiabatic evolution, this extension implies
$A_{\rm i}\p{\eta > \eta_{\rm c}} = A_{\rm i}(\eta_{\rm c})$. The boundary
between these extended zones is denoted by the dashed black line in panel (a) of
Fig.~\ref{fig:eq_1contours}, where no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/1_areas.png}
    \caption{The solid lines show the fractional areas of each of the zones
    $A_{\rm i}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward}. The colored
    dotted lines correspond to small $\eta$ approximations used in
    Appendix~\ref{s:ad_approx}. The colored dashed lines for $\eta > \eta_{\rm
    c}$ are the effective values of $A_{\rm II}, A_{\rm III}$ for $\eta >
    \eta_{\rm c}$, denoting the points that would flow into either area under
    adiabatic decrease of $\eta$ from $\eta > \eta_{\rm c}$ (see the text). The
    vertical black dashed lines correspond to $\eta = \eta_{\rm c}$
    [Eq.\eqref{eq:etac}] and the values of $\eta$ for which $A_{\rm II}$ is
    maximized ($\eta_{\rm \max, II}$) and for which $A_{\rm III}$ is minimized
    ($\eta_{\rm \min, III}$, Eq.~\eqref{eq:eta_minIII}).}\label{fig:eq_areas}
\end{figure}

@@

In this appendix, we study the evolution of the planetary obliquity $\theta$ when
the parameter $\eta$ [or the disk mass $M_{\rm d}$; see Eqs.~\eqref{eq:eta}
and~\eqref{eq:deta_dt}] decreases sufficiently slowly that the evolution is
adiabatic. Strictly speaking, this requires the disk evolution timescale $t_{\rm
d}$ [Eq.~\eqref{eq:dmd_dt}] be much larger than all timesccales of the dynamical
system governed by the Hamiltonian [Eq.~\eqref{eq:H}]. This is of course not
possible in all cases, as the motion along the separatrix has an infinite
period.

@@

The above calculation is performed for initial condition $\theta_{\rm i}$, but,
as discussed earlier in the text, the initial spin orientation is more
appropriately described by $\theta_{\rm sd, i}$. The correct way to predict
final obliquities for such initial conditions using Eq.~\eqref{eq:nonad_q_f} is
somewhat indirect but yields good agreement with numerical results. We develop
the procedure for two cases: (i) for $\theta_{\rm sd, i} = 0$, then (ii) for
nonzero $\theta_{\rm sd, i}$. To simplify the discussion below, we consider the
limit where $\eta_{\rm i} \to \infty$, and CS2 and $\uv{l}_d$ coincide. The
nonadiabatic evolution is then described by:
\begin{itemize}
    \item When $\theta_{\rm sd, i} = 0$, the initial condition is just CS2, so
        Eq.~\eqref{eq:nonad_q_f} can be directly applied setting $\theta_{\rm i}
        = I$. This yields good agreement with numerical results (see
        Figs.~\ref{fig:nonad_3_scan} and~\ref{fig:nonad_3_scan_20}). We denote
        this final obliquity
        \begin{equation}
            \theta_{\rm f} = \sqrt{\frac{2\pi}{\epsilon}}\frac{\sin (2I)}{2}.
                \label{eq:theta_2f}
        \end{equation}
        The azimuthal angle of the spin vector does not change and remains $\phi
        = \pi$. Figures~\ref{fig:nonad_3_scan} and~\ref{fig:nonad_3_scan_20}
        show the final obliquity as function of $\epsilon$ for $\theta_{\rm
        sd,i} = 0$ and $I = 5$, $20^\circ$. We see that the agreement between
        the numerical results and Eq.~\eqref{eq:nonad_q_f} is excellent,
        particularly for the $I = 5^\circ$ case. The agreement when $I =
        20^\circ$ is slightly poorer as $\theta_{\rm i}$ is larger and hence
        less negligible in Eq.~\eqref{eq:nonad_q_f}. For $\epsilon \ll
        \epsilon_{\rm c}$, we find $\theta_f \simeq 90^\circ$, in agreement with
        the result of adiabatic evolution (see Fig.~\ref{fig:ad_ensemble}).

    \item When $\theta_{\rm sd, i}$ is permitted to be nonzero, the set of
        possible initial conditions forms a ring of angular width $\theta_{\rm
        sd, i}$ about CS2. We find that this set of initial conditions
        approximately evolves under nonadiabatic evolution to a ring of angular
        width $\theta_{\rm sd, i}$ about the $\theta_{\rm sd, i} = 0$ prediction
        given by Eq.~\eqref{eq:theta_2f}. The range of possible final
        obliquities is the range of $\theta$ spanned by this ring of possible
        outcomes. This can be solved in closed form, yielding generalization of
        Eq.~\eqref{eq:theta_2f}
        \begin{equation}
            \abs{\sqrt{\frac{2\pi}{\epsilon}}\frac{\sin \p{2I}}{2}
                - \theta_{\rm sd, i}}
            \leq \theta_{\rm f}
            \leq \sqrt{\frac{2\pi}{\epsilon}}\frac{\sin\p{2I}}{2}
                + \theta_{\rm sd, i} .\label{eq:nonad_q_f_dist}
        \end{equation}
        The absolute value in the lower bound is a result of the ring of final
        conditions crossing the North pole. Fig.~\ref{fig:nonad_3_ensemble}
        shows the numerical result of $\theta_f$ vs $\theta_{\rm sd,i}$ for $I =
        5^\circ$ and $\epsilon = 0.3$. We see that Eq.~\eqref{eq:nonad_q_f_dist}
        provides good lower and upper bounds of the final obliquity for
        $\theta_{\rm sd,i} \lesssim 45^\circ$.
\end{itemize}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/3scan.png}
    \caption{Final obliquity $\theta_{\rm  f}$ as a function of $\epsilon$ for
    $\theta_{\rm sd,i} = 0$ and $I = 5^\circ$. The shaded area, bordered by the
    black line, corresponds to the adiabatic regime estimated by
    Eq.~\eqref{eq:ad_constr}. The blue dots are numerical results, and the red
    dashed line corresponds to Eq.~\eqref{eq:nonad_q_f}, which is in good
    agreement with numerical results for $\epsilon > \epsilon_{\rm c} \approx
    0.1$ (the nonadiabatic regime); $\theta_{\rm f} \simeq  90^\circ$ in the
    adiabatic regime ($\epsilon\ll \epsilon_{\rm c}$).}\label{fig:nonad_3_scan}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/3scan_20.png}
    \caption{Same as Fig.~\ref{fig:nonad_3_scan} but for $I=20^\circ$. The right
    shaded area is where $\theta_{\rm f}$ predicted by Eq.~\eqref{eq:nonad_q_f}
    violates the assumption $\theta_{\rm f} \gg \theta_{\rm
    i}$.}\label{fig:nonad_3_scan_20}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/3_ensemble_05_05.png}
    \caption{$\theta_{\rm  f}$ vs $\theta_{\rm sd, i}$ for $I=5^\circ$
    and $\epsilon = 0.3$ (firmly in the nonadiabatic regime). The blue
    dots represent numerical results, and the two red lines show the
    analytical lower and upper limits given by
    Eq.~\eqref{eq:nonad_q_f_dist}.}\label{fig:nonad_3_ensemble}
\end{figure}

@@

In this section, we linearize the equations of motion near each CS\@. We
determine the stability of each CS and the local libration frequencies or growth
rates of the CSs.

\subsection{Canonical Equations of Motion and Solutions}\label{ss:canonical}

We adopt spherical coordinate system where $\uv{l} =
\uv{z}$ and $\theta, \phi$ are the polar and azimuthal angle of $\uv{s}$. We
choose $\uv{l}_{\rm d}$ at coordinates $\theta = I, \phi = \pi$ (see
Figs.~\ref{fig:cs_vecs} and~\ref{fig:cs_locs}). We use the convention
$\theta \in [0, \pi)$% chktex 9
and $\phi \in [0, 2\pi)$.% chktex 9

The equations of motion in $\p{\phi, \cos \theta}$ follow by
applying Hamilton's equations to the Hamiltonian [Eq.~\eqref{eq:H}]:
\begin{subequations}\label{se:H_eom}
    \begin{align}
        \rd{\phi}{t} = \pd{\mathcal{H}}{(\cos\theta)}
            &= -\cos\theta + \eta\p{\cos I + \sin I \cot \theta \cos \phi},
                \label{seq:H_eom_phi_t}\\
        \rd{(\cos \theta)}{t} = -\pd{\mathcal{H}}{\phi}
            &= -\eta \sin I \sin \theta \sin \phi.
                \label{seq:H_eom_mu_t}
    \end{align}
\end{subequations}
These agree with Eq.~\eqref{eq:dsdt_base}.
% While these equations are numerically stiff owing to the $\cot\theta$ term, they
% are the most intuitive description for analytical work.

% We next develop some approximate forms for the CS obliquities that will be
% useful for guiding later discussion. Denote $\theta_{\rm cs}$ the obliquity of
% some Cassini State (fixed point of Eqs.~\eqref{se:H_eom}).
% \begin{enumerate}
%     \item $\eta \ll 1, \cos \theta_{\rm cs} \ll 1$ --- This is the limiting case
%         for CS2 and CS4 when $\eta \ll \eta_{\rm c}$. We examine
%         Eqs.~\eqref{seq:H_eom_phi_t} and approximate $\cot \theta \approx \cos
%         \theta$ to find
%         \begin{equation}
%             \cos \theta_{\rm cs} \approx \frac{\eta \cos I}
%                 {1 \mp \eta \sin I}.
%         \end{equation}
%         The two signs correspond to choices of $\cos \phi$. Each sign
%         corresponds to one of CS2 and CS4.

%     \item $\sin \theta_{\rm cs} \ll 1$ --- The $\eta \ll 1, \eta \gg 1$ cases
%         can be solved together, and are the limiting cases both for CS1 and CS3
%         when $\eta \ll \eta_{\rm c}$, and for CS2 and CS3 when $\eta \gg
%         \eta_{\rm c}$. It proves easiest to rewrite Eqs.~\eqref{seq:H_eom_phi_t}
%         as
%         \begin{equation}
%             0 = \cos \theta_{\rm cs}\p{-1 + 2\eta
%                 \frac{\sin \p{I \pm \theta_{\rm cs}}}{\sin \p{2\theta_{\rm
%                     cs}}}}.\label{eq:rewritten_dphi}
%         \end{equation}
%         The $\pm$ choice again comes from choice of $\cos \phi$. Then assuming
%         $\theta_{\rm cs}, I \ll 1$, we obtain $\frac{I \pm
%         \theta_{\rm cs}}{2\theta_{\rm cs}} = \frac{1}{2\eta}$ or
%         \begin{equation}
%             \theta_{\rm cs} = \frac{\eta I}{1 \mp \eta}.
%         \end{equation}

%         In the limit $\eta \ll \eta_{\rm c}$, these two solutions describe CS1
%         and CS3, while in the limit $\eta \gg \eta_{\rm c}$ these two solutions
%         describe CS2 and CS3.
% \end{enumerate}
% Again, the $\theta_{\rm cs}$ values derived here do not follow the same
% convention as Fig.~\ref{fig:cs_locs} due to our different $\theta, \phi$
% convention.

The CSs satisfy $\dot{\phi} = \dot{\theta} = 0$. For convenience, we give
approximate solutions for the CSs in the limits $\eta \ll 1$ and $\eta \gg 1$.
For $\eta \ll 1$:
\begin{itemize}
    \item CS1: $\phi_1 = 0$, $\theta_1 \simeq \eta \sin I$.
    \item CS2: $\phi_2 = \pi$, $\theta_2 \simeq \pi / 2 + \eta \cos I$.
    \item CS3: $\phi_3 = 0$, $\theta_3 \simeq \pi - \eta \sin I$.
    \item CS4: $\phi_4 = 0$, $\theta_4 \simeq \pi / 2 - \eta \cos I$.
\end{itemize}
For $\eta \gg 1$, only CS2 and CS3 exist and are given by:
\begin{itemize}
    \item CS2: $\phi_2 = \pi$, $\theta_2 \simeq I - \sin I \cos I / \eta$.
    \item CS3: $\phi_3 = 0$, $\theta_3 \simeq \pi - I + \sin I \cos I / \eta$.
\end{itemize}

@@

In this appendix, we will use approximations valid for small $\eta$ to derive
the explicit analytic expressions for the final obliquities at small
$\theta_{\rm sd, i}$ and the associated probabilities for the II $\to$ I and II
$\to$ III tracks. These are the only possible tracks for small $\eta$.

We first seek a simple parameterization for the separatrix, the level curve of
the Hamiltonian intersecting the unstable equilibrium CS4. Points along the
separatrix, parameterized by $\p{\phi, \theta_{\rm sep}(\phi)}$, satisfy
$\mathcal{H}\p{\phi, \theta_{\rm sep}(\phi)} = \mathcal{H}\p{\phi_4, \theta_4}$
where $\phi_4$ and $\theta_4$ are given in Appendix~\ref{ss:canonical}. We
obtain two solutions for $\theta_{\rm sep}$, given to leading order in $\eta$
by:
\begin{equation}
    \cos \theta_{\rm sep}(\phi) \approx \cos \theta_4 \pm
        \sqrt{2\eta \sin I\p{1 - \cos \phi}}.
\end{equation}
These two solutions parameterize the two legs of the separatrix. Integration of
the phase area enclosed by the separatrix yields then
\begin{equation}
    A_{\rm II}(\eta) \approx 16\sqrt{\eta \sin I}.\label{eq:a_approx}
\end{equation}
We can now compute the final obliquities and their associated probabilities for
each track as follows:
\begin{enumerate}
    \item For a given $\theta_{\rm sd, i}$, we know that if $\eta \to \infty$
        then the trajectory executes simple libration about $\uv{l}_{\rm d}$,
        and so $A = 2\pi\p{1 - \cos \theta_{\rm sd, i}} \approx \pi \theta_{\rm
        sd, i}^2$. This then implies $\eta_\star$ must be the solution to
        $A_{\rm II}(\eta_\star) = A$, or
        \begin{equation}
            \eta_\star \approx \p{\frac{2\pi\p{1 - \cos \theta_{\rm sd,i}}}{
                        16}}^2 \frac{1}{\sin I}
                    \approx \p{\frac{\pi \theta_{\rm sd,
                    i}^2}{16}}^2\frac{1}{\sin I}.
        \end{equation}
